---
title: 'STA 360 Lab 2: R review'
author: "Isaac Fan"
date: "1/29/2021"
output: pdf_document
---

```{r setup, message=F, warning=F, echo=F}
#
require(tidyverse)
require(rstanarm)
require(magrittr)
#
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(fig.align = 'center')
```

# Preliminaries

Please turn in a PDF of this Rmd file on Sakai by **Friday, February 5th at 11:59 PM**. Exercises 5 and 6 will be graded for completion. No other exercises will be graded, so it is your choice to do them. 


### Exercise 5 (For completion)
(a) Generate $500$ samples from a Beta distribution with shape parameter $[a, b] = [0.5, 0.5]$ and store the samples in a variable called `W`.

```{r betasamp}
W <- rbeta(500, .5, .5, ncp = 0)

```

(b) Generate $1000$ samples from a binomial distribution with parameters $[n, p] = [100, 0.2]$ and store them in a variable called `B` (hint: read the documentation to make sure you're inputting the parameters in the correct spots). Calculate the mean of `B` and compare this to the mean of this binomial distribution ($n p$).

```{r binomsamp}
B <- rbinom(1000, 100, .2)

```

```{r mean}
mean(W)
mean(B)
```

The mean of the Binomial distribution is significantly higher than the mean of the Beta distribution by roughly 40 times.

# Plotting

When making plots in R, you have two main options: (1) the base R plotting function `plot` and (2) the package `ggplot2`. 

The base R plotting functions are nice for quick, simple visualizations of data. Here are some examples:

```{r example plots}
norm_samples <- rnorm(10000)
#
par(mfrow = c(2, 2)) # Set the number of rows and columns for display panels
#
hist(norm_samples,
     main = "Base R histogram", 
     xlab = "x", ylab = "Count")
#
plot(x = norm_samples, y = pnorm(norm_samples), 
     main = "Base R scatterplot", 
     xlab = "x", ylab = "Phi(x)")
#
boxplot(norm_samples, 
        main = "Base R boxplot", 
        ylab = "x")
#
plot(density(norm_samples),
     main = "Base R density", 
     xlab = "x", ylab = "Density")
```

Plotting with `ggplot2` is generally a little bit easier when working with data in large tables. It has a gallery of built-in themes, and there are many extensions that make producing complicated visualizations relatively straightforward.



***
### Exercise 6 (For Completion)

The following line of code produces a time series of length 550 and stores it in the variable `x`. 

```{r plotfun}
#
x0 <- rt(1, df = 5)
xprev <- x0
x <- rep(0, 550)
x[1] <- xprev
for(i in 1:549){
  x[i+1] <- 0.8*xprev + rnorm(1)
  xprev <- x[i+1]
}
```

(a) Make a histogram of `x`. 

```{r ex6a}
hist(x)

```


(b) Make a plot of `x` (on the y-axis) against time (on the x-axis). 


```{r ex6b}
plot(x = time(x), y = x, 
     main = "x vs Time", 
     xlab = "Time", ylab = "x")

```

***

# R tutorials and resources

For more information on the R programming language please refer to some or all of the following resources:

- [R for Data Science](https://r4ds.had.co.nz)

- [Advanced R](https://adv-r.hadley.nz/)

- [ggplot2](https://ggplot2.tidyverse.org)

- [RStan documentation](https://mc-stan.org/rstan/)

- [ColorBrewer](http://colorbrewer2.org)
