---
title: "HW2"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r theta}
theta <- c(0,.1,.2,.3,.4,.5,.6,.7,.8,.9,1)
```



```{r prob}
#P(Y=57 | theta)
prob <- c(0,0,0,0,0,0,0,0,0,0,0)
for (i in 1:11)
{
prob[i] <- theta[i]^57 * (1-theta[i])^43
}
prob
```

## Part B

```{r plot}
plot(x = theta, y = prob, ylab = "P(Y = 57 | theta)")
```



```{r bayes}
sum(prob)
# P(theta | Y = 57)
prob_c <- c(0,0,0,0,0,0,0,0,0,0,0)
for (i in 1:11)
{
prob_c[i] <- prob[i] * (1/11) / sum(prob)
}
prob_c
```

## Part C

```{r part c}
plot(x = theta, y = prob_c, ylab = "P(theta | Y = 57)")
```


## Part D
```{r part d}
theta_cont <- seq(from = 0, to = 1, by = .001)
prob_cont <- c()
for (i in theta_cont)
{
p <- i^57 * (1-i)^43
prob_cont <- append(prob_cont, p)
}
prob_cont
```


```{r}
plot(x = theta_cont, y = prob_cont, ylab = "P(Y = 57 | theta)")
```

## Part E

```{r part e}
x = seq(0,1, by = .001)
plot(x, dbeta(x, 58, 44), ylab="density", type ="l", col=4)
```

They all appear to be similar beta distributions but scaled differently. Plot C is a posterior with the prior being uniform probability of each theta occuring. Plot B has no prior information given to us. Plot D and plot E should be the exact same beta distribution but scaled differently. Plot D uses the prior of uniform pdf for theta, whereas E is the true posterior distribution.